Tutorial: fitting mixed-effects and log-Gaussian mixture models on text-writing data

### 2023-02-08 tutorial/tutorial.Rmd
Update tutorial.rmd which can not be compiled

---
title: "Tutorial: fitting mixed mixture models on text writing data"
author: "Jens Roeser"
date: "Compiled `r Sys.Date()`"
bibliography: ref.bib
csl: apa.csl
link-citations: yes
output:
  rmdformats::readthedown:
    lightbox: true
    gallery: false
    highlight: "pygments"
    toc_depth: 3
    use_bookdown: false

@@

```{r setup, include=FALSE}
options(scipen = 999) # Turn off scientific notation
knitr::opts_knit$set(width=90)
knitr::opts_chunk$set(fig.width = 8, 
                      fig.height = 4.5,
                      echo = TRUE,
                      comment = NA, 
                      warning = FALSE,
                      message = FALSE)
theme_set(theme_bw() + theme(legend.justification = "top"))
```

@@

```{r}
iterations <- 30000       # No of iterations
warmup <- iterations/2    # Warmup samples to be discarded
nchains <- ncores <- 3  # No of chains / cores used (one chain per core)
```

@@

## Model comparison

@@

## Mixture model app

The data must be transformed into a list to feed them into Stan. The information required from the data is determined in the Stan code and has to do with what we need to estimate the model parameters and how the model is implemented. The exert from the Stan code below shows which information is expected as input (e.g. int is integer), how they are named, and what the smallest (i.e. lower) and largest possible value (i.e. upper) are. The full Stan code can be found at the end of this document.

@@

```{r eval = F}
# Fit model
fit_molg <- sampling(molg, 
                     data = data_list,
                     iter = iterations,
                     warmup = warmup,
                     chains = nchains, 
                     cores = ncores,
                     pars = pars, # Parameters to keep.
                     save_warmup = FALSE, # Don't save the warmup samples.
                     seed = 365, 
                     # Improve sampling efficiency
                     control = list(adapt_delta = 0.96,  # default: 0.9
                                    max_treedepth = 14)) # default: 10
```

```{r echo =F}
#sampler_params <- get_sampler_params(m, inc_warmup = FALSE)
#sampler_params_chain1 <- sampler_params[[3]]#colnames(sampler_params_chain1)
#sapply(sampler_params, function(x) mean(x[, "accept_stat__"]))
#sapply(sampler_params, function(x) max(x[, "treedepth__"]))
#inits <- get_inits(m)
#inits_chain1 <- inits[[1]]
#print(inits_chain1)
```

@@

```{r}
summary(fit_molg, pars=pars)$summary %>% 
  as.data.frame() %>% 
  rownames_to_column("parameter") %>% 
  select(parameter, Rhat)
```

@@

## Difference between copy-task components

We can calculate the differences between the copy-task components for each parameter value. This is giving an indication of whether fluent typing ($\beta$) is influenced by the copy task, maybe the size of the disfluency slowdown ($\delta$) is impacted by the task, or the disfluency probability ($\theta$).

To determine these differences we change the data format above and create a variable that indicates the parameter and one column for the corresponding values for each copy-task component. The last line calculates the difference between the consonants task and the LF-bigrams task for each of the three parameter values.

```{r}
posterior_by_component <- posterior_in_msecs %>%
  pivot_longer(beta:theta, names_to = "parameter") %>%
  group_by(component) %>% mutate(id = row_number()) %>%
  pivot_wider(names_from = component, values_from = value) %>%
  select(-id) %>% # drop id column
  mutate(diff = Consonants - `LF bigrams`,  # calculate the difference between copy-task components
         parameter = paste0(parameter, "[diff]"))

posterior_by_component
```

@@

```{r}
posterior_by_component %>%
  group_by(parameter) %>%                      # group by parameters
  summarise(mean = mean(diff),                 # mean difference
            lower = quantile(diff, .025),      # 2.5% lower bound of difference
            upper = quantile(diff, .975),      # 97.5% upper bound of difference
            p = mean(diff < 0)) %>%            # prob that difference is negative
  mutate(across(where(is.numeric), round, 2))  # round after 2nd decimal place
```

```{r echo=F}
ps <- posterior_by_component %>%
  group_by(parameter) %>%
  summarise(mean = mean(diff),
            lower = quantile(diff, .025),
            upper = quantile(diff, .975),
            p = mean(diff < 0)) %>% 
  mutate(across(where(is.numeric), round, 2),
         parameter = gsub("\\[diff\\]", "", parameter)) 
```

These differences can be viewed in histograms. The vertical line indicates a difference between tasks corresponding to a value of zero. The area of the histogram to the left of this line corresponds to the probability `p` to observe a value smaller than zero as in the output above. We observe difference between copy-task components for all three parameters. Fluent typing is `r round(ps[ps$parameter == "beta",]$mean)` msecs (95% PI: `r round(ps[ps$parameter == "beta",]$lower)`, `r round(ps[ps$parameter == "beta",]$upper)`) slower for the consonants task, and hesitations are `r round(ps[ps$parameter == "delta",]$mean)` msecs (95% PI: `r round(ps[ps$parameter == "delta",]$lower)`, `r round(ps[ps$parameter == "delta",]$upper)`) longer and `r ps[ps$parameter == "theta",]$mean` ($\approx$ `r round(ps[ps$parameter == "theta",]$mean*100)`%; 95% PI: `r ps[ps$parameter == "theta",]$lower`, `r ps[ps$parameter == "theta",]$upper`) times more likely. Interestingly fluent typing (i.e. planning and executing motor plans) is slower in the consonant task even after modeling hesitations as a separate process.

```{r}
ggplot(posterior_by_component, aes(x = diff)) +
  geom_histogram(alpha = .5) +
  geom_vline(xintercept = 0, colour = "red", linetype = "dashed", size = .5) +
  facet_wrap(~parameter, scales = "free", labeller = label_parsed) +
  labs(x = "Difference between copy-task components")
```

In practice we rarely believe that only an effect of exactly null is indicating that there is no difference between our groups. Instead there are values that are non-different from null; for example a slowdown of 1 msecs is meaningless in most, if not all,  contexts. To take this into account we can determine the *region of practical equivalence* [ROPE, @kruschke2010believe; @kruschke2011bayesian; @kruschke2014doing]. We can define the ROPE as the range of values that is _a priori_ considered non-different from a null effect. The ROPE value indicates to what extent the posterior can not rule out a negligible effect. A meaningful difference between groups (or effects more generally) should have a small proportion of posterior samples within the ROPE.

To calculate the ROPE for out differences between the two copy-task components we can use the `rope` function from the `bayestestR` package.^[To install `bayestestR` run `install.packages("bayestestR")`.] The `rope` function requires as input a vector with the posterior difference. We can calculate a vector with posterior difference for each of the three parameters, indicated as $\beta_\text{diff}$, $\delta_\text{diff}$, and $\theta_\text{diff}$.

```{r}
# Fluent typing
beta_diff <- filter(posterior_by_component, parameter == "beta[diff]") %>% pull(diff)

# Slowdown for hesitant typing
delta_diff <- filter(posterior_by_component, parameter == "delta[diff]") %>% pull(diff)

# Proportion of hesitations
theta_diff <- filter(posterior_by_component, parameter == "theta[diff]") %>% pull(diff)
```

@@

```{r}
library(bayestestR)
rope(beta_diff, range = c(-10, 10))
```

The ROPE for fluent typing speed contains `r round(pull(rope(beta_diff, range = c(-10, 10)))*100,2)`% of the posterior samples. In other words, there is a `r round(pull(rope(beta_diff, range = c(-10, 10)))*100,2)`% probability that the posterior difference $\beta_\text{diff}$ is between -10 and 10 msecs which we considered negligible.

For disfluencies we might want to consider a larger range of values as negligible differences between two tasks (also because the variability in disfluencies is larger than the variability for fluent key transitions). Even if we define a ROPE of -250 through 250 msecs, there are no posterior samples inside the ROPE indicating that the difference in typing hesitations are related to difficulty that arise prior to processing on the motor level.

```{r}
rope(delta_diff, range = c(-250, 250))
```

As for the disfluency probably, we need to bear in mind that estimates for $\theta$ are bound between 0 and 1. Hence, the difference $\theta_\text{diff}$ cannot ever exceed -1 and 1. If we consider a difference of 5% or smaller as negligible, we can define the ROPE as ranging from -.05 through .05. We observe that `r round(pull(rope(theta_diff, range = c(-.05, .05)))*100,2)`% of the posterior difference $\theta_\text{diff}$ is within the ROPE. In other words, a large proportion of posterior samples is supporting a meaningful difference between the two copy-task components. We observe that hesitations are largely more frequent in the consonants task.

```{r}
rope(theta_diff, range = c(-.05, .05))
```

@@

In some contexts we would like to obtain by-participant estimates of the average fluent typing-speed of a participant after accounting for disfluencies (i.e. $\beta_s$); by-participant disfluency information can be estimated the proportion of disfluencies (i.e. $\theta_s$). These estimates inform us about the fluent typing speed but also about the prevalence of typing hesitations.

As before we can extract the posterior from the model `m`. This time we use the parameters that stored by-participant estimates `beta_s` and `theta_s` corresponding to the population estimates `beta` and `theta`.

```{r}
pars <- c("beta_s", "theta_s")
posterior_ppts <- as.data.frame(m, pars = pars) %>% as_tibble()
names(posterior_ppts)[1:10]
```

This posterior tibble has the format "$<\text{parameter name}>\text{_s}[<\text{component id}>, <\text{participant id}>]$" where indices indicate the copy-task component and the participant identifier of the parameters. The following code is converting the `beta`s to msecs, creates a long format with columns `parameter` with the names above as levels and a `value` column with their respective values. The `parameter` column is then separated into `parameter` (levels: `beta`, `theta`), `component` (levels: `1`, `2`), and participant (an index for each participant) using "," to separate the three variables. To use "," as separator we replace "_s" with "," in the line before. Copy-task component names were changed as before. A preview is below.

```{r}
posterior_ppts_long <- posterior_ppts %>%
  # use exp() on everything starting with "beta"
  mutate(across(starts_with("beta"), exp)) %>%             
  pivot_longer(everything(), names_to = "parameter") %>%
  # replace "_s" with "," 
  mutate(parameter = gsub("_s", ",", parameter)) %>%
  # separate into parameter, component, ppt using "," as separator
  separate(parameter, into = c("parameter", "component", "participant")) %>% 
  mutate(component = recode(component, `1` = "Consonants", `2` = "LF bigrams")) %>%
  group_by(parameter) %>% mutate(id = row_number()) %>%
  pivot_wider(names_from = parameter, values_from = value) %>% 
  select(-id) # drop id

@@

We can then summarise by-participant estimates for each component and parameter with the most probable parameter value and the 95% PI as before.

```{r}
posterior_ppts_long_summary <- posterior_ppts_long %>%
  pivot_longer(beta:theta, names_to = "parameter", values_to = "value") %>%
  group_by(parameter, component, participant) %>%
  summarise(mean = mean(value),
            lower = quantile(value, .025),
            upper = quantile(value, .975))

@@

To get an idea how the by-participant estimates look like, we extracted participants 1 through 5 for visualisation. From this subset we see that, although we observed overall larger population estimates for the consonants task, some participants do not show a difference in fluent typing speed and a similar frequency of hesitations (e.g. participant 2) or even slightly more hesitations in the LF-bigrams task (e.g. participant 1).

```{r }
posterior_ppts_long_summary %>%
  filter(participant %in% 1:5) %>% # only participants 1 through 5
  ggplot(aes(x = participant, 
             y = mean, ymin = lower, ymax = upper,
            colour = component)) +
  geom_pointrange(position = position_dodge(.75)) +
  facet_wrap(~parameter, scales = "free", labeller = label_parsed) +
  coord_flip() +
  labs(x = "Participant id", y = "Parameter estimates") +
  scale_colour_brewer("Copy-task\ncomponent", palette = "Dark2")
```

@@

```{r}
posterior_ppts_long_diff_summary <- posterior_ppts_long %>% group_by(component) %>% 
  mutate(id = row_number()) %>%
  pivot_wider(names_from = component, 
              values_from = c(beta, theta)) %>%
  pivot_longer(-c(participant, id),
             names_to = c("parameter", ".value"), 
             names_pattern = "(.*)_(.*)") %>%
  mutate(diff = Consonants - `LF bigrams`,
         parameter = paste0(parameter, "[diff]")) %>% 
  select(participant, parameter, diff) %>%
  group_by(parameter, participant) %>%
  summarise(mean = mean(diff),
            lower = quantile(diff, .025),
            upper = quantile(diff, .975))

@@

For simplicity we show the differences for the first 10 participants only. This calculation and visualisation allows researchers to identify participants that do or do not differ in terms of fluent typing speed or hesitation frequency, in our example for tasks that do and do not involve lexical information.

```{r }
posterior_ppts_long_diff_summary %>%
  filter(participant %in% 1:10) %>% # only participants 1 through 10
  ggplot(aes(x = factor(as.numeric(participant)), 
             y = mean, ymin = lower, ymax = upper)) +
  geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
  geom_pointrange(position = position_dodge(.75)) +
  facet_wrap(~parameter, scales = "free", labeller = label_parsed) +
  coord_flip() +
  labs(x = "Participant id", y = "Difference between copy-task components") +
  scale_colour_brewer("Copy-task\ncomponent", palette = "Dark2")
```

@@

```{r echo = F}
cat(readLines("MoG.stan"), sep = "\n")
```

>>>>

---
title: "Tutorial: fitting mixed-effects and log-Gaussian mixture models on text-writing data"
author: "Jens Roeser"
date: "Compiled `r Sys.Date()`"
bibliography: ref.bib
csl: apa.csl
link-citations: yes
output:
  rmdformats::readthedown:
    lightbox: true
    gallery: false
    highlight: "pygments"
    toc_depth: 3
    use_bookdown: false

@@

```{r setup, include=FALSE}
# Turn off scientific notation
options(scipen = 999)

# Chunk defaults
knitr::opts_knit$set(width=90)
knitr::opts_chunk$set(fig.width = 8, 
                      fig.height = 4.5,
                      size = "small",
                      echo = TRUE,
                      comment = NA, 
                      warning = FALSE,
                      message = FALSE)

# Plotting defaults
theme_set(theme_bw() + 
            theme(legend.justification = "top",
                  panel.grid = element_blank()))
```

@@

```{r}
iterations <- 10000      # No of iterations
warmup <- iterations / 2 # Warmup samples to be discarded
nchains <- ncores <- 3   # No of chains / cores used (one chain per core)
```

@@

## TODO Model comparison

ADD here info from slides on how to calculate elpd

@@

## TODO Mixture model simulation

## Model fit

The data must be transformed into a list to feed them into Stan. The information required from the data is determined in the Stan code and has to do with what we need to estimate the model parameters and how the model is implemented. The exert from the Stan code below shows which information is expected as input (e.g. int is integer), how they are named, and what the smallest (i.e. lower) and largest possible value (i.e. upper) are. The full Stan code can be found at the end of this tutorial.

@@

```{r eval = F}
# Fit model
fit_molg <- sampling(molg, 
                     data = data_list,
                     iter = iterations,
                     warmup = warmup,
                     chains = nchains, 
                     cores = ncores,
                     pars = pars, # Parameters to keep.
                     save_warmup = FALSE, # Don't save the warmup samples.
                     seed = 365, 
                     # Improve sampling efficiency
                     control = list(adapt_delta = 0.94,  # default: 0.9
                                    max_treedepth = 12)) # default: 10
```

```{r echo = F}
#sampler_params <- get_sampler_params(m, inc_warmup = FALSE)
#sampler_params_chain1 <- sampler_params[[3]]#colnames(sampler_params_chain1)
#sapply(sampler_params, function(x) mean(x[, "accept_stat__"]))
#sapply(sampler_params, function(x) max(x[, "treedepth__"]))
#inits <- get_inits(m)
#inits_chain1 <- inits[[1]]
#print(inits_chain1)
```

@@

```{r}
summary(fit_molg, pars=pars)$summary %>% 
  as.data.frame() %>% 
  rownames_to_column("parameter") %>% 
  mutate(across(Rhat, round, 3)) %>% 
  select(parameter, Rhat)
```

@@

## Difference between transition locations

We can calculate the differences between the transition locations for each parameter. This is giving an indication of whether fluent typing ($\beta$) is influenced by linguistic edges, maybe the size of the disfluency slowdown ($\delta$), or the disfluency probability ($\theta$).

To determine these differences we change the data format above and create a variable that indicates the parameter and one column for the corresponding values for each transition location. The last line calculates the difference between (1) transitions before sentences and words, and (2) transitions before and within words for each of the three parameters.

```{r}
posterior_diffs <- posterior_in_msecs %>%
  pivot_longer(beta:theta, names_to = "parameter") %>%
  group_by(location) %>% 
  mutate(id = row_number()) %>%
  pivot_wider(names_from = location, values_from = value) %>%
  select(-id) %>% # drop id column
  mutate(diff_sent = `before-sentence` - `before-word`,
         diff_word = `before-word` - `within-word`,
         parameter = paste0(parameter, "[diff]")) %>% 
  select(parameter, starts_with("diff"))

posterior_diffs
```

@@

```{r}
# Create some summary functions
lower <- function(x) quantile(x, .025)
upper <- function(x) quantile(x, .975)
p_diff <- function(x) mean(x < 0)

# Summarise differences
posterior_diffs %>%
  pivot_longer(starts_with("diff")) %>% 
  group_by(parameter, name) %>%      # group by parameters, comparison
  summarise(across(value, 
            list(mean = base::mean,  # mean difference
                 lower = lower,      # 2.5% lower bound of difference
                 upper = upper,      # 97.5% upper bound of difference
                 p_diff = p_diff),   # prob that difference is negative
            .names = "{.fn}")) %>% 
  ungroup() %>% 
  mutate(across(where(is.numeric), round, 2)) %>% # round after 2nd decimal place
  arrange(name)
```

```{r echo=F}
ps <- posterior_diffs %>%
  pivot_longer(-parameter) %>% 
  group_by(parameter, name) %>%
  summarise(across(value, 
            list(mean = base::mean,  
                 lower = lower,      
                 upper = upper,      
                 p_diff = p_diff),   
            .names = "{.fn}")) %>% 
  ungroup() %>% 
  mutate(across(where(is.numeric), round, 2),
         parameter = gsub("\\[diff\\]", "", parameter)) 
```

These differences can be viewed in histograms. The vertical line indicates a difference between tasks corresponding to a value of zero. The area of the histogram to the left of this line corresponds to the probability `p` to observe a value smaller than zero as in the output above.

```{r}
posterior_diffs %>% 
  pivot_longer(starts_with("diff")) %>% 
  mutate(across(name, recode, diff_sent = "before-sentence vs before-word",
                              diff_word = "before-word vs within word")) %>% 
  ggplot(aes(x = value, colour = name, fill = name)) +
  geom_histogram(position = "identity", alpha = .25) +
  facet_wrap(~parameter, scales = "free", labeller = label_parsed) +
  scale_fill_brewer("Difference", palette = "Dark2") +
  scale_color_brewer("Difference", palette = "Dark2") +
  geom_vline(xintercept = 0, colour = "red", linetype = "dashed", size = .5) +
  labs(x = "Difference between transition locations") +
  theme(legend.position = "top")
```

We observe the following differences that can be written up in a report like this: We observe evidence for a higher probability to pause before sentences which is `r ps %>% filter(parameter == "theta", name == "diff_sent") %>% pull(mean)` (95% PI: `r ps %>% filter(parameter == "theta", name == "diff_sent") %>% pull(lower)`, `r ps %>% filter(parameter == "theta", name == "diff_sent") %>% pull(upper)`) times more likely than before words, which was is `r ps %>% filter(parameter == "theta", name == "diff_word") %>% pull(mean)` (95% PI: `r ps %>% filter(parameter == "theta", name == "diff_word") %>% pull(lower)`, `r ps %>% filter(parameter == "theta", name == "diff_word") %>% pull(upper)`) more likely than within words.

Similarly, differences were found for the hesitation difference. Hesitations before sentences are `r ps %>% filter(parameter == "delta", name == "diff_sent") %>% pull(mean)` msecs (95% PI: `r ps %>% filter(parameter == "delta", name == "diff_sent") %>% pull(lower)`, `r ps %>% filter(parameter == "delta", name == "diff_sent") %>% pull(upper)`) longer than before words, which were `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(mean)` msecs (95% PI: `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(lower)`, `r ps %>% filter(parameter == "delta", name == "diff_word") %>% pull(upper)`) longer than within words.

Interestingly though fluent transitions (i.e. planning and executing motor plans) were only faster within words by `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(mean)` msecs (95% PI: `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(lower)`, `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(upper)`) compared to before-word transitions but there was no difference for before-sentence transitions compared to before-word transitions ($\hat{\beta}$ = `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(mean)` msecs, 95% PI: `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(lower)`, `r ps %>% filter(parameter == "beta", name == "diff_word") %>% pull(upper)`).

## ROPE

In practice, we rarely believe that only an effect of exactly null is indicating that there is no difference between our groups (which is what a null hypothesis assumes). Instead there are values that are non-different from null; for example a slowdown of 1 msecs is meaningless in most, if not all, contexts. To take this into account we can determine the *region of practical equivalence* [ROPE, @kruschke2010believe; @kruschke2011bayesian; @kruschke2014doing]. We can define the ROPE as the range of values that is _a priori_ considered non-different from a null effect. The ROPE value indicates to what extent the posterior can not rule out a negligible effect. A meaningful difference between groups (or effects more generally) should have a small proportion of posterior samples within the ROPE.

To calculate the ROPE for out differences between two transition locations, we can use the `rope` function from the `bayestestR` package.^[To install `bayestestR` run `install.packages("bayestestR")`.] The `rope` function requires as input a vector with the posterior difference. We can create such a vector from our posterior differences. Let's focus on $\beta_\text{diff}$, the average typing speed for fluent transitions.

```{r}
# Extract vector of differences
# For sentence vs word
beta_diff_sent <- filter(posterior_diffs, parameter == "beta[diff]") %>% pull(diff_sent)

# For before word vs within word
beta_diff_word <- filter(posterior_diffs, parameter == "beta[diff]") %>% pull(diff_word)
```

@@

```{r}
library(bayestestR)
rope(beta_diff_sent, range = c(-10, 10))
```

The ROPE for fluent typing speed for the difference between before-sentence and before-word transitions contains `r round(pull(rope(beta_diff_sent, range = c(-10, 10)))*100,2)`% of the posterior samples. In other words, there is a `r round(pull(rope(beta_diff_sent, range = c(-10, 10)))*100,2)`% probability that the posterior difference $\beta_\text{diff}$ is between -10 and 10 msecs which we considered negligible.

```{r}
rope(beta_diff_word, range = c(-10, 10))
```

However, the ROPE for fluent typing speed for the difference between before-word and within-word transitions contains `r round(pull(rope(beta_diff_word, range = c(-10, 10)))*100,2)`% of the posterior samples. In other words, there is a `r round(pull(rope(beta_diff_word, range = c(-10, 10)))*100,2)`% probability that the posterior difference $\beta_\text{diff}$ is between -10 and 10 msecs which we considered negligible.

Let's see what happens if we determine a wider ROPE for negligible differences. What happens if we define a ROPE of -50 through 50 msecs for the difference between before-word and within-word transitions.

```{r}
rope(beta_diff_word, range = c(-50, 50))
```

The ROPE contains `r round(pull(rope(beta_diff_word, range = c(-50, 50)))*100,2)`% of the posterior samples. In other words, there is a `r round(pull(rope(beta_diff_word, range = c(-50, 50)))*100,2)`% probability that the posterior difference $\beta_\text{diff}$ is between -50 and 50 msecs which we considered negligible.

## TODO Bayes Factor

Add info from Savage Dickey Bayes Factor slide

## Note on using more than a single IV

@@

In some contexts we would like to obtain by-participant estimates of the probability of a participant to exhibit a disfluency (i.e. $\theta_s$). These estimates inform us about the prevalence of typing hesitations.

As before we can extract the posterior from the model `fit_molg`. This time we use the parameter that stored by-participant estimates `prob_s` (which is the inverse-logit of the `theta_s` which we didn't store in the model fit) corresponding to the population estimate `theta`.

```{r}
posterior_ppts <- as.data.frame(fit_molg, pars = 'prob_s') %>% as_tibble()
names(posterior_ppts)[1:10]
```

This posterior tibble has the format

| "$<\text{parameter name}>\text{_s}[<\text{transition location id}>, <\text{participant id}>]$"

where indices indicate the transition-location id and the participant identifier of the parameters. The following code creates a long format with columns `name` with the names above as levels and a `value` column with their respective values (both `name` and `value` are `pivot_longer` defaults). The `name` column is then separated into `name` (levels: `prob`), `location` (levels: `1`, `2`, `3`), and participant (an index for each participant) using "," to separate the three variables.

```{r}
posterior_ppts_long <- posterior_ppts %>%
  pivot_longer(everything()) %>%
  # separate into location and ppt using "," as separator
  separate(name, into = c("location", "participant"), sep = ",") %>% 
  # extract number from new variables and recode location
  mutate(across(c(location, participant), parse_number),
         across(location, recode, 
                `1` = "before-sentence",
                `2` = "before-word",
                `3` = "within-word")) %>%
  # create different columns for location
  group_by(location) %>% 
  mutate(id = row_number()) %>%
  pivot_wider(names_from = location, values_from = value) %>% 
  select(-id) # drop id

@@

We can then summarise by-participant estimates for each transition location with the most probable parameter value and the 95% PI as before.

```{r}
posterior_ppts_long_summary <- posterior_ppts_long %>%
  pivot_longer(-participant, names_to = "location") %>%
  group_by(location, participant) %>%
  summarise(across(value, list(mean = base::mean, 
                               lower = lower, 
                               upper = upper), .names = '{.fn}')) %>% 
  ungroup()

@@

For all 10 participants we see the estimate probability of typing hesitations by transition location. We observe in general that hesitations are more frequent before sentences and words than within words. However, participants vary to the extend they hesitate before words and sentences suggesting that some participant plan sentences more globally and others plan sentences word-by-word (incrementally).

```{r}
posterior_ppts_long_summary %>%
  ggplot(aes(x = participant, 
             y = mean, ymin = lower, ymax = upper,
            colour = location,
            shape = location)) +
  geom_pointrange(position = position_dodge(.75)) +
  labs(x = "Participant id", 
       y = "Parameter estimates with 95% PIs") +
  scale_colour_brewer("Transition\nlocation", palette = "Dark2") +
  scale_shape_manual("Transition\nlocation", values = c(21, 3, 8)) +
  scale_x_continuous(breaks = 1:10) +
  theme(legend.position = 'top')
```

@@

```{r}
posterior_ppts_long_diff_summary <- posterior_ppts_long %>% 
  mutate(diff_sent = `before-sentence` - `before-word`,
         diff_word = `before-word` - `within-word`) %>% 
  select(participant, starts_with('diff')) %>%
  pivot_longer(-participant, 
               names_to = 'difference', 
               values_to = 'diff') %>% 
  group_by(participant, difference) %>%
  summarise(across(diff, list(mean = base::mean, 
                               lower = lower, 
                               upper = upper), .names = '{.fn}')) %>% 
  ungroup()

@@

This calculation and visualisation allows researchers to identify participants that do or do not differ in terms of hesitation frequency, in our example depending on transition locations.

```{r }
posterior_ppts_long_diff_summary %>%
  mutate(across(difference, recode, 
                diff_sent = "Comparison: before-sentence vs before-word",
                diff_word = "Comparison: before-word vs within-word")) %>% 
  ggplot(aes(x = factor(as.numeric(participant)), 
             y = mean, ymin = lower, ymax = upper)) +
  geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
  geom_pointrange(position = position_dodge(.75)) +
  facet_wrap(~difference) +
  coord_flip() +
  labs(x = "Participant id", 
       y = "Difference between transition locations") 
```

@@

```{r echo = F}
cat(readLines("stan/molg.stan"), sep = "\n")
```
